Orienting the causal relationship between imprecisely measured traits using genetic instruments

### 2017-03-03 docs/manuscript.tex
updated

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Inferring the existence and direction of causal associations in the face of measurement error},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Inferring the existence and direction of causal associations in the face
of measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{24 February 2017}

@@

Mediation-based approaches employ genetic instruments (typically single
nucleotide polymorphisms, SNPs) to orient the causal direction between
the exposure and the outcome. If a SNP is associated with an exposure,
and the exposure is associated with some outcome, then it logically
follows that the indirect influence of the SNP on the outcome will
disappear when conditioning on the exposure. Here, the exposure
`mediates' the association between the SNP and the outcome, providing
information about the causal influence of the exposure on the outcome.
This forms the basis of a number of methods such as genetical genomics
(17), the regression-based causal inference test (CIT) (4,18), a
structural equation modelling (SEM) implementation in the NEO software
(5), and various other methods including Bayesian approaches (6). They
have been employed by a number of recent publications that make causal
inferences in large scale `omics datasets (6,19--23).

@@

MR has some important advantages over the mediation-based approaches.
The mediation-based approaches require that the exposure, outcome and
instrumental variables are all measured in the same data, whereas recent
extensions to MR circumvent this requirement, allowing causal inference
to be drawn when exposure variables and outcome variables are measured
in different samples (28). This has the crucial advantage of improving
statistical power by allowing analysis in much larger sample sizes, and
dramatically expands the breadth of possible phenotypic relationships
that can be evaluated (26). Additionally, when MR assumptions are
satisfied the method is robust to there being measurement error in the
exposure variable (29). Indeed instrumental variable (IV) analysis was
in part initially introduced as a correction for measurement error in
the exposure (30), whereas it has been noted that both classic
mediation-based analyses (13,14,31,32) and mediation-based methods that
use instrumental variables (33,34) are prone to be unreliable in its
presence.

Using theory and simulations we show how non-differential measurement
error in phenotypes can lead to unreliable causal inference in the
mediation-based CIT method. We then present an extension to MR that
allows researchers to ascertain the causal direction of an association
even when the biology of the instruments are not fully understood, and
also a metric to evaluate the sensitivity of the result of this
extension to measurement error. Together these extensions improve the
utility of MR in cases where mediation based methods might have
otherwise been used preferentially. Finally, we apply this method to
infer the direction of causation between DNA methylation levels and gene
expression levels. Our analyses highlight that because these different
causal inference techniques have varying strengths and weaknesses,
triangulation of evidence from as many sources as possible should be
practiced in causal inference (35).

@@

Measurement error of an exposure can be modeled as a transformation of
the true value that leads to the observed value, \(x_o = f(x)\). For
example, following Pierce and VanderWeele (29) we can define

@@

where \(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the
residual of \(y\) after adjusting for \(x\), where \(x\) is assumed to
mediate the association between the SNP and the outcome. If all four
tests reject the null hypothesis then it is inferred that \(x\) causes
\(y\). The CIT measures the strength of causality by generating an
omnibus p-value, \(p_{CIT}\), which is simply the largest (least
extreme) p-value of the four tests, the intuition being that causal
inference is only as strong as the weakest link in the chain of tests.

@@

For the purposes of compiling simulation results we use an arbitrary
\(\alpha = 0.05\) value, though we stress that for real analyses it is
not good practice to rely on p-values for making causal inference, nor
is it reliable to depend on arbitrary significance thresholds (36).

@@

Two stage least squares (2SLS) is a commonly used technique for
performing MR when the exposure, outcome and instrument data are all
available in the same sample. A p-value for this test, \(p_{MR}\), was
obtained using the R package \(R/systemfit\) (37).

@@

To infer the direction of causality it is desirable to know which of the
variables, \(x\) or \(y\), is being directly influenced by the
instrument \(g\). This can be achieved by assessing which of the two
variables has the biggest absolute correlation with \(g\) (Appendix 2),
formalised by testing for a difference in the correlations \(\rho_{gx}\)
and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
within a population (38). It is calculated as

@@

An advantage of using the Steiger test in the two sample context is that
it can compare correlations in independent samples where sample sizes
are different. Steiger test statistics were calculated using the
\emph{r.test} function in the R package \emph{R/psych} (39).

@@

Two sample MR (28) was performed using the summary statistics for
genetic influences on gene expression and DNA methylation. To do this we
obtained a list of 458 gene expression - DNA methylation associations as
reported in Shakhbazov et al (40). These were filtered to be located on
the same chromosome, have robust correlations after correcting for
multiple testing, and to share a SNP that had a robust cis-acting effect
on both the DNA methylation probe and the gene expression probe. Because
only summary statistics were available (effect, standard error, effect
allele, sample size, p-values) for the instrumental SNP on the
methylation and gene expression levels, the Steiger test of two
independent correlations was used to infer the direction of causality
for each of the associations. The Wald ratio test was then used to
estimate the causal effect size for the estimated direction for each
association.

All analysis was performed using the R programming language (41) and
code is made available at \href{}{github url to go here} and implemented
in the MR-Base (\url{http://wwww.mrbase.org}) platform (26).

@@

We performed two sample MR (28) for each association in the direction of
causality inferred by the Stieger test. We observed that the sign of the
MR estimate was generally in the same direction as the Pearson
correlation coefficient reported by Shakhbazov et al (40) (Figure 6b).
There was a moderate correlation between the absolute magnitudes of the
causal correlation and the observational Pearson correlation (r = 0.45).
Together these inferences suggest that even for relatively low level
associations it is important to use causal inference methods over
observational associations to infer causal effect sizes.

We also observed that for associations where methylation caused gene
expression the causal effect was more likely to be negative than for the
associations where gene expression caused methylation (OR = 0.61 (95\%
CI 0.36 - 1.03), Figure 6c), suggesting that reducing methylation levels
at a controlling CpG typically leads to increased gene expression
levels, consistent with expectation (42).

@@

Researchers are often confronted with the problem of making causal
inferences using a statistical framework on observational data. In the
epidemiological literature issues of measurement error in mediation
analysis is relatively well explored (43). Our analysis extends this to
related methods such as CIT that are widely used in predominantly 'omic
data. These methods are indeed liable to the same issues as standard
mediation based analysis, and specifically we show that as measurement
error in the exposure variable increases, CIT is likely to have reduced
statistical power, and liable to infer the wrong direction of causality.
We also demonstrate that, though unintuitive, increasing sample size
does not resolve the issue, rather it leads to more extreme p-values for
the model that predicts the wrong direction of causality.

@@

We demonstrate this new method by evaluating the causal relationships of
458 known associations between DNA methylation and gene expression
levels using summary level data. The causal direction is heavily
influenced by how much measurement error is present in the different
assaying platforms. For example, if DNA methylation measures typically
have higher measuremet error than gene expression measures then our
analysis suggests that DNA methylation levels would be more often the
causal factor in the association. Indeed, previous studies which have
evaluated measurement error in these platforms do support this position
(44,45), though making strong conclusions for this analysis is difficult
because measurement error is likely to be study specific.

In our simulations we focused on the simple case of a single instrument
in a single sample setting with a view to making a fair comparison
between MR and the various mediation-based methods available. We note
that it is straightforward to extend the MR Steiger approach to multiple
instruments, requiring only that the total variance explained by all
instruments be calculated under the assumption that they are
independent. Multiple instruments can indeed help to circumvent some of
the problems described here in other ways also, for example by
evaluating the proportionality of the SNP-exposure and SNP-outcome
effects (16). Additionally, if there is at least one instrument for each
trait then bi-directional MR can offer solutions to inferring the causal
direction (46). We restricted the simulations to evaluating the causal
inference between quantitative traits, but it is possible that the
analysis could be extended to binary traits by using the genetic
variance explained on the liability scale, taking into account the
population prevalence (47).

In this work we assumed that pleiotropy (the influence of the instrument
on the outcome through a mechanism other than the exposure) was not
present. Recent method developments in MR (24,25) have focused on
accounting for the issues that horizontal pleiotropy can introduce when
multiple instruments are available, but how they perform in the presence
of measurement error remains to be explored. An important advantage that
MR confers over most mediation based analysis is that it can be
performed in two samples, which can considerably improve power and
expand the scope of analysis. However, whether there is a substantive
difference in two sample MR versus one sample MR in how measurement
error has an effect is not yet fully understood. We have also assumed no
measurement error in the genetic instrument, which is not unreasonable
given the strict QC protocols that ensure high quality genotype data is
available to most studies. We have restricted the scope to only
exploring non-differential measurement error and avoided the
complications incurred if measurement error in the exposure and outcome
is correlated. We have also not addressed other issues pertaining to
instrumental variables, such as exposure missspecification, where the
instrument is used for a particular variable due to that variable's
non-causal correlation with the true exposure. However our analysis goes
beyond many previous explorations of measurement error by assessing the
impacts of both imprecision (noise) and linear transformations of the
true variable on causal inference.

Mediation based network approaches, that go beyond analyses of two
variables, are very well established (34) and have a number of
extensions that make them valuable tools, including for example network
construction. But because they are predicated on the basic underlying
principles of mediation they are liable to suffer from the same issues
of measurement error. Recent advances in MR methodology, for example
applying MR to genetical genomics (48), multivariate MR (49) and
mediation through MR (50--52) may offer more robust alternatives for
these more complicated problems.

@@

\begin{itemize}
\tightlist
\item
  other types of measurement error, e.g.~differential; exposure
  missspecification;
\item
  using multiple instruments
\end{itemize}

@@

\hypertarget{ref-Pierce2012}{}
29. Pierce BL, VanderWeele TJ. The effect of non-differential
measurement error on bias, precision and power in Mendelian
randomization studies. International Journal of Epidemiology
{[}Internet{]}. 2012 Oct;41(5):1383--93. Available from:
\url{https://academic.oup.com/ije/article-lookup/doi/10.1093/ije/dys141}

\hypertarget{ref-Ashenfelter1994}{}
30. Ashenfelter O, Krueger AB. Estimates of the Economic Return to
Schooling from a New Sample of Twins. The American Economic Review.
1994;84(5):1157--73.

\hypertarget{ref-Nagarajan2013}{}
31. Nagarajan R, Scutari M. Impact of noise on molecular network
inference. PloS one {[}Internet{]}. 2013 Jan;8(12):e80735. Available
from:
\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3855153\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3855153\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}

\hypertarget{ref-Shpitser2010}{}
32. Shpitser I, VanderWeele T, Robins J. On the validity of covariate
adjustment for estimating causal effects. Proceedings of the Twenty
Sixth Conference on Uncertainty in Artificial Intelligence (UAI-10).
2010;527--36.

\hypertarget{ref-Wang2015}{}
33. Wang L, Michoel T. Detection of regulator genes and eQTLs in gene
networks. arXiv {[}Internet{]}. 2015 Dec;arXiv:1512. Available from:
\url{http://arxiv.org/abs/1512.05574}

\hypertarget{ref-Lagani2015}{}
34. Lagani V, Triantafillou S, Ball G, Tegner J, Tsamardinos I.
Probabilistic Computational Causal Discovery for Systems Biology. In:
Uncertainty in biology: A computational modeling approach
{[}Internet{]}. Springer; 2015. p. 47. Available from:
\href{https://books.google.com/books?id=8SLUCgAAQBAJ\%7B/\&\%7Dpgis=1}{https://books.google.com/books?id=8SLUCgAAQBAJ\{\textbackslash{}\&\}pgis=1}

\hypertarget{ref-Lawlor2017}{}
35. Lawlor DA, Tilling K, Davey Smith G. Triangulation in aetiological
epidemiology. International Journal of Epidemiology {[}Internet{]}. 2017
Jan;19(R1):dyw314. Available from:
\url{https://academic.oup.com/ije/article-lookup/doi/10.1093/ije/dyw314}

\hypertarget{ref-Sterne2001}{}
36. Sterne JAC, Smith GD. Sifting the evidence---what's wrong with
significance tests? BMJ. 2001;322(7280):226--31.

\hypertarget{ref-Henningsen2007}{}
37. Henningsen A, Hamann JD. systemfit : A Package for Estimating
Systems of Simultaneous Equations in R. Journal of Statistical Software
{[}Internet{]}. 2007 Dec;23(4):1--40. Available from:
\url{https://www.jstatsoft.org/index.php/jss/article/view/v023i04/v23i04.pdf}

\hypertarget{ref-Steiger1980}{}
38. Steiger JH. Tests for comparing elements of a correlation matrix.
Psychological Bulletin. 1980;87(2):245--51.

\hypertarget{ref-Revelle2015}{}
39. Revelle W. psych: Procedures for Psychological, Psychometric, and
Personality Research {[}Internet{]}. Evanston, Illinois: Northwestern
University; 2015. Available from:
\url{http://cran.r-project.org/package=psych}

\hypertarget{ref-Shakhbazov2016}{}
40. Shakhbazov K, Powell JE, Hemani G, Henders AK, Martin NG, Visscher
PM, et al. Shared genetic control of expression and methylation in
peripheral blood. BMC genomics {[}Internet{]}. 2016 Jan;17(1):278.
Available from:
\url{http://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-016-2498-4}

\hypertarget{ref-RCoreTeam2015}{}
41. R Core Team. R: A Language and Environment for Statistical Computing
{[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
2015. Available from: \url{https://www.r-project.org/}

\hypertarget{ref-Bird2002}{}
42. Bird A. DNA methylation patterns and epigenetic memory. Genes \&
development {[}Internet{]}. 2002 Jan;16(1):6--21. Available from:
\url{http://genesdev.cshlp.org/content/16/1/6.long}

\hypertarget{ref-Cole2014}{}
43. Cole DA, Preacher KJ. Manifest Variable Path Analysis: Potentially
Serious and Misleading Consequences Due to Uncorrected Measurement
Error. Psychological Methods. 2014;19(2):300--15.

\hypertarget{ref-Bose2014}{}
44. Bose M, Wu C, Pankow JS, Demerath EW, Bressler J, Fornage M, et al.
Evaluation of microarray-based DNA methylation measurement using
technical replicates: the Atherosclerosis Risk In Communities (ARIC)
Study. BMC Bioinformatics {[}Internet{]}. 2014;15(1):312. Available
from: \url{http://www.biomedcentral.com/1471-2105/15/312}

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Orienting the causal relationship between imprecisely measured traits using genetic instruments},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Orienting the causal relationship between imprecisely measured traits
using genetic instruments}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{03 March 2017}

@@

Mediation-based approaches employ genetic instruments (typically single
nucleotide polymorphisms, SNPs) to orient the causal direction between
the exposure and the outcome. If a SNP is associated with an exposure,
and the exposure is associated with some outcome, then it logically
follows that in this simple three-variable scenario the indirect
influence of the SNP on the outcome will disappear when conditioning on
the exposure. Here, the exposure `mediates' the association between the
SNP and the outcome, providing information about the causal influence of
the exposure on the outcome. This forms the basis of a number of methods
such as genetical genomics (17), the regression-based causal inference
test (CIT) (4,18), a structural equation modelling (SEM) implementation
in the NEO software (5), and various other methods including Bayesian
approaches (6). They have been employed by a number of recent
publications that make causal inferences in large scale `omics datasets
(6,19--23).

@@

MR has some important advantages over the mediation-based approaches.
The mediation-based approaches require that the exposure, outcome and
instrumental variables are all measured in the same data, whereas recent
extensions to MR circumvent this requirement, allowing causal inference
to be drawn when exposure variables and outcome variables are measured
in different samples (28). This has the crucial advantage of improving
statistical power by allowing analysis in much larger sample sizes, and
dramatically expands the breadth of possible phenotypic relationships
that can be evaluated (26). It is important to note that if there are
any confounders influencing both the exposure and outcome, the
mediation-based approach of adjusting the outcome for the exposure to
nullify the association between the SNP and the outcome will not work.
This is because adjusting the outcome by the exposure induces a collider
effect between the SNP and outcome (29), and the in order to fully
abbrogate this association one must also adjust for all (hidden or
otherwise) confounders. MR does not suffer from this problem because it
does not test for association through adjustment. Additionally, when MR
assumptions are satisfied the method is robust to there being
measurement error in the exposure variable (30). Indeed instrumental
variable (IV) analysis was in part initially introduced as a correction
for measurement error in the exposure (31), whereas it has been noted
that both classic mediation-based analyses (13,14,32,33) and
mediation-based methods that use instrumental variables (34,35) are
prone to be unreliable in its presence.

Using theory and simulations we show how non-differential measurement
error in phenotypes can lead to unreliable causal inference in the
mediation-based CIT method. We then present an extension to MR that
allows researchers to ascertain the causal direction of an association
even when the biology of the instruments are not fully understood, and
also a metric to evaluate the sensitivity of the result of this
extension to measurement error. Together these extensions improve the
utility of MR in cases where mediation based methods might have
otherwise been used preferentially. Finally, we apply this method to
infer the direction of causation between DNA methylation levels and gene
expression levels. Our analyses highlight that because these different
causal inference techniques have varying strengths and weaknesses,
triangulation of evidence from as many sources as possible should be
practiced in causal inference (36).

@@

Measurement error of an exposure can be modeled as a transformation of
the true value that leads to the observed value, \(x_o = f(x)\). For
example, following Pierce and VanderWeele (30) we can define

@@

where \(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the
residual of \(y\) after adjusting for \(x\), where \(x\) is assumed to
mediate the association between the SNP and the outcome. The 4th
condition is formulated as an equivalence testing problem that is
estimated using simulations, comparing the estimate against from the
data against empirically obtained estimates for simulated variables
where the independence model is true (full details are given in (4)). We
note here that this approach is liable to fail, even when there is a
true causal relationship, when confounders are present due to inducing
collider effects.

If all four tests reject the null hypothesis then it is inferred that
\(x\) causes \(y\). The CIT measures the strength of causality by
generating an omnibus p-value, \(p_{CIT}\), which is simply the largest
(least extreme) p-value of the four tests, the intuition being that
causal inference is only as strong as the weakest link in the chain of
tests.

@@

For the purposes of compiling simulation results we use an arbitrary
\(\alpha = 0.05\) value, though we stress that for real analyses it is
not good practice to rely on p-values for making causal inference, nor
is it reliable to depend on arbitrary significance thresholds (37).

@@

Two stage least squares (2SLS) is a commonly used technique for
performing MR when the exposure, outcome and instrument data are all
available in the same sample. A p-value for this test, \(p_{MR}\), was
obtained using the R package \(R/systemfit\) (38).

@@

To infer the direction of causality it is desirable to know which of the
variables, \(x\) or \(y\), is being directly influenced by the
instrument \(g\). This can be achieved by assessing which of the two
variables has the biggest absolute correlation with \(g\) (Appendix 2),
formalised by testing for a difference in the correlations \(\rho_{gx}\)
and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
within a population (39). It is calculated as

@@

An advantage of using the Steiger test in the two sample context is that
it can compare correlations in independent samples where sample sizes
are different. Steiger test statistics were calculated using the
\emph{r.test} function in the R package \emph{R/psych} (40).

@@

Two sample MR (28) was performed using the summary statistics for
genetic influences on gene expression and DNA methylation. To do this we
obtained a list of 458 gene expression - DNA methylation associations as
reported in Shakhbazov et al (41). These were filtered to be located on
the same chromosome, have robust correlations after correcting for
multiple testing, and to share a SNP that had a robust cis-acting effect
on both the DNA methylation probe and the gene expression probe. Because
only summary statistics were available (effect, standard error, effect
allele, sample size, p-values) for the instrumental SNP on the
methylation and gene expression levels, the Steiger test of two
independent correlations was used to infer the direction of causality
for each of the associations. The Wald ratio test was then used to
estimate the causal effect size for the estimated direction for each
association.

All analysis was performed using the R programming language (42) and
code is made available at \href{}{github url to go here} and implemented
in the MR-Base (\url{http://wwww.mrbase.org}) platform (26).

@@

We performed two sample MR (28) for each association in the direction of
causality inferred by the Stieger test. We observed that the sign of the
MR estimate was generally in the same direction as the Pearson
correlation coefficient reported by Shakhbazov et al (41) (Figure 6b).
There was a moderate correlation between the absolute magnitudes of the
causal correlation and the observational Pearson correlation (r = 0.45).
Together these inferences suggest that even for relatively low level
associations it is important to use causal inference methods over
observational associations to infer causal effect sizes.

We also observed that for associations where methylation caused gene
expression the causal effect was more likely to be negative than for the
associations where gene expression caused methylation (OR = 0.61 (95\%
CI 0.36 - 1.03), Figure 6c), suggesting that reducing methylation levels
at a controlling CpG typically leads to increased gene expression
levels, consistent with expectation (43).

@@

Researchers are often confronted with the problem of making causal
inferences using a statistical framework on observational data. In the
epidemiological literature issues of measurement error in mediation
analysis is relatively well explored (44). Our analysis extends this to
related methods such as CIT that are widely used in predominantly 'omic
data. These methods are indeed liable to the same issues as standard
mediation based analysis, and specifically we show that as measurement
error in the exposure variable increases, CIT is likely to have reduced
statistical power, and liable to infer the wrong direction of causality.
We also demonstrate that, though unintuitive, increasing sample size
does not resolve the issue, rather it leads to more extreme p-values for
the model that predicts the wrong direction of causality.

@@

We demonstrate this new method by evaluating the causal relationships of
458 known associations between DNA methylation and gene expression
levels using summary level data. The causal direction is heavily
influenced by how much measurement error is present in the different
assaying platforms. For example, if DNA methylation measures typically
have higher measuremet error than gene expression measures then our
analysis suggests that DNA methylation levels would be more often the
causal factor in the association. Indeed, previous studies which have
evaluated measurement error in these platforms do support this position
(45,46), though making strong conclusions for this analysis is difficult
because measurement error is likely to be study specific.

In our simulations we focused on the simple case of a single instrument
in a single sample setting with a view to making a fair comparison
between MR and the various mediation-based methods available. We note
that it is straightforward to extend the MR Steiger approach to multiple
instruments, requiring only that the total variance explained by all
instruments be calculated under the assumption that they are
independent. Multiple instruments can indeed help to circumvent some of
the problems described here in other ways also, for example by
evaluating the proportionality of the SNP-exposure and SNP-outcome
effects (16). Additionally, if there is at least one instrument for each
trait then bi-directional MR can offer solutions to inferring the causal
direction (47). We restricted the simulations to evaluating the causal
inference between quantitative traits, but it is possible that the
analysis could be extended to binary traits by using the genetic
variance explained on the liability scale, taking into account the
population prevalence (48). However, our analysis goes beyond many
previous explorations of measurement error by assessing the impacts of
both imprecision (noise) and linear transformations of the true variable
on causal inference.

In this work we assumed that pleiotropy (the influence of the instrument
on the outcome through a mechanism other than the exposure) was not
present. Recent method developments in MR (24,25) have focused on
accounting for the issues that horizontal pleiotropy can introduce when
multiple instruments are available, but how they perform in the presence
of measurement error remains to be explored. An important advantage that
MR confers over most mediation based analysis is that it can be
performed in two samples, which can considerably improve power and
expand the scope of analysis. However, whether there is a substantive
difference in two sample MR versus one sample MR in how measurement
error has an effect is not yet fully understood. We have also assumed no
measurement error in the genetic instrument, which is not unreasonable
given the strict QC protocols that ensure high quality genotype data is
available to most studies. We have restricted the scope to only
exploring non-differential measurement error and avoided the
complications incurred if measurement error in the exposure and outcome
is correlated. We have also not addressed other issues pertaining to
instrumental variables which are relevant to the question of
instrument-exposure specification. One such problem is exposure
misspecification, for example an instrument could associate with several
closely related putative outcomes, with only one of them actually having
a causal effect on the outcome. This problem has shown to be the case
for SNPs influencing different lipid fractions, for example (49,50).

Mediation based network approaches, that go beyond analyses of two
variables, are very well established (35) and have a number of
extensions that make them valuable tools, including for example network
construction. But because they are predicated on the basic underlying
principles of mediation they are liable to suffer from the same issues
of measurement error. Recent advances in MR methodology, for example
applying MR to genetical genomics (51), multivariate MR (50) and
mediation through MR (52--54) may offer more robust alternatives for
these more complicated problems.

@@

@@

\hypertarget{ref-Hernan2004}{}
29. Hernán MA, Hernández-Díaz S, Robins JM. A structural approach to
selection bias. Epidemiology (Cambridge, Mass) {[}Internet{]}. 2004
Sep;15(5):615--25. Available from:
\url{http://www.ncbi.nlm.nih.gov/pubmed/15308962}

\hypertarget{ref-Pierce2012}{}
30. Pierce BL, VanderWeele TJ. The effect of non-differential
measurement error on bias, precision and power in Mendelian
randomization studies. International Journal of Epidemiology
{[}Internet{]}. 2012 Oct;41(5):1383--93. Available from:
\url{https://academic.oup.com/ije/article-lookup/doi/10.1093/ije/dys141}

\hypertarget{ref-Ashenfelter1994}{}
31. Ashenfelter O, Krueger AB. Estimates of the Economic Return to
Schooling from a New Sample of Twins. The American Economic Review.
1994;84(5):1157--73.

\hypertarget{ref-Nagarajan2013}{}
32. Nagarajan R, Scutari M. Impact of noise on molecular network
inference. PloS one {[}Internet{]}. 2013 Jan;8(12):e80735. Available
from:
\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3855153\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3855153\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}

\hypertarget{ref-Shpitser2010}{}
33. Shpitser I, VanderWeele T, Robins J. On the validity of covariate
adjustment for estimating causal effects. Proceedings of the Twenty
Sixth Conference on Uncertainty in Artificial Intelligence (UAI-10).
2010;527--36.

\hypertarget{ref-Wang2015}{}
34. Wang L, Michoel T. Detection of regulator genes and eQTLs in gene
networks. arXiv {[}Internet{]}. 2015 Dec;arXiv:1512. Available from:
\url{http://arxiv.org/abs/1512.05574}

\hypertarget{ref-Lagani2015}{}
35. Lagani V, Triantafillou S, Ball G, Tegner J, Tsamardinos I.
Probabilistic Computational Causal Discovery for Systems Biology. In:
Uncertainty in biology: A computational modeling approach
{[}Internet{]}. Springer; 2015. p. 47. Available from:
\href{https://books.google.com/books?id=8SLUCgAAQBAJ\%7B/\&\%7Dpgis=1}{https://books.google.com/books?id=8SLUCgAAQBAJ\{\textbackslash{}\&\}pgis=1}

\hypertarget{ref-Lawlor2017}{}
36. Lawlor DA, Tilling K, Davey Smith G. Triangulation in aetiological
epidemiology. International Journal of Epidemiology {[}Internet{]}. 2017
Jan;19(R1):dyw314. Available from:
\url{https://academic.oup.com/ije/article-lookup/doi/10.1093/ije/dyw314}

\hypertarget{ref-Sterne2001}{}
37. Sterne JAC, Smith GD. Sifting the evidence---what's wrong with
significance tests? BMJ. 2001;322(7280):226--31.

\hypertarget{ref-Henningsen2007}{}
38. Henningsen A, Hamann JD. systemfit : A Package for Estimating
Systems of Simultaneous Equations in R. Journal of Statistical Software
{[}Internet{]}. 2007 Dec;23(4):1--40. Available from:
\url{https://www.jstatsoft.org/index.php/jss/article/view/v023i04/v23i04.pdf}

\hypertarget{ref-Steiger1980}{}
39. Steiger JH. Tests for comparing elements of a correlation matrix.
Psychological Bulletin. 1980;87(2):245--51.

\hypertarget{ref-Revelle2015}{}
40. Revelle W. psych: Procedures for Psychological, Psychometric, and
Personality Research {[}Internet{]}. Evanston, Illinois: Northwestern
University; 2015. Available from:
\url{http://cran.r-project.org/package=psych}

\hypertarget{ref-Shakhbazov2016}{}
41. Shakhbazov K, Powell JE, Hemani G, Henders AK, Martin NG, Visscher
PM, et al. Shared genetic control of expression and methylation in
peripheral blood. BMC genomics {[}Internet{]}. 2016 Jan;17(1):278.
Available from:
\url{http://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-016-2498-4}

\hypertarget{ref-RCoreTeam2015}{}
42. R Core Team. R: A Language and Environment for Statistical Computing
{[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
2015. Available from: \url{https://www.r-project.org/}

\hypertarget{ref-Bird2002}{}
43. Bird A. DNA methylation patterns and epigenetic memory. Genes \&
development {[}Internet{]}. 2002 Jan;16(1):6--21. Available from:
\url{http://genesdev.cshlp.org/content/16/1/6.long}

\hypertarget{ref-Cole2014}{}
44. Cole DA, Preacher KJ. Manifest Variable Path Analysis: Potentially
Serious and Misleading Consequences Due to Uncorrected Measurement
Error. Psychological Methods. 2014;19(2):300--15.

\hypertarget{ref-Bose2014}{}
45. Bose M, Wu C, Pankow JS, Demerath EW, Bressler J, Fornage M, et al.
Evaluation of microarray-based DNA methylation measurement using
technical replicates: the Atherosclerosis Risk In Communities (ARIC)
Study. BMC Bioinformatics {[}Internet{]}. 2014;15(1):312. Available
from: \url{http://www.biomedcentral.com/1471-2105/15/312}

\hypertarget{ref-Bryant2011}{}
46. Bryant PA, Smyth GK, Robins-Browne R, Curtis N, Novak J, Sladek R,
et al. Technical Variability Is Greater than Biological Variability in a
Microarray Experiment but Both Are Outweighed by Changes Induced by
Stimulation. Khodursky AB, editor. PLoS ONE {[}Internet{]}. 2011
May;6(5):e19556. Available from:
\url{http://dx.plos.org/10.1371/journal.pone.0019556}
